Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-26 numerical.tex
editing numerical results, redid mmsn application for 50au and corrected espilon to include mu dependence

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_iso}
  \caption{Eigenvalues of the VSI in a neutrally-stratified,
    nearly-vertically isothermal disk  ($\gamma=\Gamma=1.011$). The
    disk parameters are $q=-1$, $p=-1.5$, $\epsilon=0.1$, and the perturbation radial
    wavenumber is $\khat=20\pi$. \label{lowfreq_eigen}
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_adia}
  \caption{Eigenvalues in for the
    VSI in a nearly vertically isothermal disk
    ($\Gamma=1.011$) evolved adiabatically with $\gamma=2$. The disk
    parameters are $q=-1$, 
    $p=-1.5$, $\epsilon=0.1$, and the perturbation radial
    wavenumber is $\khat=20\pi$. \label{lowfreq_eigen_adia}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvectorvz_adia}
  \caption{Fundamental VSI mode in a nearly vertically
    isothermal disk ($\Gamma=1.011$) evolved adiabatically
    with $\gamma=2$. This eigenfunction 
    corresponds to the top-left eigenvalue displayed in 
    Fig. \ref{lowfreq_eigen_adia} (largest $|\nu|$).  
    The  real (imaginary) part of $\delta v_z$ are shown as solid
    (dotted) lines. 
    \label{lowfreq_eigenfunc_adia}
  }
\end{figure}

@@

\subsection{Other adiabatic indices}
In Fig. \ref{adia_growth_num} we plot the eigenfrequencies of the
fundamental VSI mode for $\gamma\in[1,2]$. To select the appropriate
eigenvalues we vary the parameters slowly away from a reference solution.

% The numerically-determined eigenfrequencies are qualitatively
% consistent with the discussion in \S\ref{adia_compare_gam1}.
There is good agreement between Fig. \ref{adia_growth_num} and
Fig. \ref{adia_growth} for $k_xH_\mathrm{iso}\gtrsim10$. There is a
mistach for smaller wavenumbers, especially in the growth
rates. Actual growth rates decrease slower as $k_xH_\mathrm{iso}\to0$
than that in Fig. \ref{adia_growth}. Disagreement at small
$k_xH_\mathrm{iso}$ is not surprising because the low-frequency
approximation used to obtain Fig. \ref{adia_growth} becomes invalid since
$|\omega|\sim\Omega_k$ for small $k_x$.

Given the number of approximations made in \S\ref{adia_compare_gam1},
Fig. \ref{adia_growth_num} compares well with
Fig. \ref{adia_growth}. The main qualitative features of the full
numerical solution in Fig. \ref{adia_growth_num} are captured by
Fig. \ref{adia_growth}: stabilization for
$k_xH_\mathrm{iso}\to0$ and $k_xH_\mathrm{iso}\to\infty$,
stabilization for increasing $\gamma$ (but ineffective at small $k_x$), 
and the shift of the most unstable wavenumber to smaller values as
$\gamma$ is increased.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_real1}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag1}
  \caption{Real frequency (top) and growth rate 
    (bottom) of the fundamental VSI mode in nearly-vertically   
    isothermal disks ($\Gamma=1.011$) with $\epsilon=0.1$ and  
    $q=-1$, subject to adiabatic perturbations
    ($\beta=10^9$). The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. This plot is to
    be compared with Fig. \ref{adia_growth}. \label{adia_growth_num}}   
\end{figure}

\subsection{Effect of thermal relaxation}
We now introduce thermal relaxation. Here we consider a slightly
different disk model with $\epsilon=0.05$ and $\gamma=1.4$, as
adopted in \cite{nelson13}. Fig. \ref{relax_growth_num} plots the
growth rates of the fundamental mode function of $k_x$. %corrugation
                                %mode 
% We first obtain the eigenvalues for the
% fundamental mode for $k_xH_\mathrm{iso}=0$, then vary parameters
% slowly to obtain eigenvalues for other values of $k_x$ or $\beta$.

As above, Fig. \ref{relax_growth_num} is largely consistent with 
Fig. \ref{relax_disp_fig} except for $k_xH_\mathrm{iso}\lesssim 1$
where the low-frequency approximation fails. The non-smoothness in the
curve for $\beta=0.063$ at $k_xH_\mathrm{iso}\sim50$ was found to be
associated with large perturbations near the vertical boundaries. This is
likely attributable to the free surface boundary condition imposed in
the numerical calculation (cf. regularity condition at infinity in
\S\ref{disp_relax}).

Fig. \ref{relax_growth_num} confirms our analytical discussion that
introducing a small but finite cooling time stabilizes the fundamental
VSI, but there is a preferable wavenumber around $k_xH_\mathrm{iso} =
O(10)$ that maximizes this effect.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_imag_bloop}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag_bloop2}
  \caption{Growth rates of the fundamental VSI mode in
    nearly-vertically isothermal disks ($\Gamma=1.011$) with
    $\epsilon=0.05$ and $q=-1$, with thermal relaxation timescale
    $\beta$. The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. \label{relax_growth_num}}   
\end{figure}

@@

\subsubsection{$N_z^2\equiv 0$}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_vnoniso1} 
  \caption{Numerically-determined growth rates of the fundamental VSI
    mode in neutrally-buoyant, vertically non-isothermal disks
    ($\Gamma=\gamma=1.4$), with $p=0$, $q=-1$ and  $\epsilon=0.05$, for
    a range of thermal relaxation timescales
    $\beta$. \label{growth_vnoniso1}}     
\end{figure}

@@

\subsubsection{$N_z^2\geq0$}

@@

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_iso}
  \caption{Eigenvalues of the VSI in a neutrally-stratified,
    nearly-vertically isothermal disk  ($\gamma=\Gamma=1.011$) with 
    $(p,q,\epsilon)=(-1.5,-1,0.1)$, and the perturbation radial
    wavenumber is $\khat=20\pi$. \label{lowfreq_eigen}
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_adia}
  \caption{Eigenvalues in for the
    VSI in a nearly vertically isothermal disk
    ($\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$, 
    evolved adiabatically with $\gamma=2$. The perturbation radial
    wavenumber is $\khat=20\pi$. \label{lowfreq_eigen_adia}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvectorvz_adia}
  \caption{Fundamental VSI mode in a nearly vertically
    isothermal disk ($\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$, 
    evolved adiabatically
    with $\gamma=2$. This eigenfunction 
    corresponds to the top-left eigenvalue displayed in 
    Fig. \ref{lowfreq_eigen_adia} (largest $|\nu|$).  
    The  real (imaginary) part of $\delta v_z$ are shown as solid
    (dotted) lines. 
    \label{lowfreq_eigenfunc_adia}
  }
\end{figure}

@@

\subsection{Other adiabatic indices}
To see the transition from neutrally-bouyant to stably stratified cases above, 
we repeated the previous calculation with $\gamma\in[1,2]$. 
We restore the vertical domain to $\zmax=10\epsilon r$. 
Fig. \ref{adia_growth_num} shows the eigenfrequencies of the
fundamental VSI mode.

%To select the appropriate
%eigenvalues we vary the parameters slowly away from a reference solution.   
% The numerically-determined eigenfrequencies are qualitatively
% consistent with the discussion in \S\ref{adia_compare_gam1}.
There is good agreement between Fig. \ref{adia_growth_num} and
Fig. \ref{adia_growth} for $\khat=k_xH_\mathrm{iso}\gtrsim10$. There is a
mismatch for smaller wavenumbers, especially in the growth
rates. Actual growth rates decrease slower as $\khat\to0$
than that in Fig. \ref{adia_growth}. Disagreement at small
$\khat$ is not unexpected because the low-frequency
approximation used to obtain Fig. \ref{adia_growth} becomes invalid since
$|\omega|\sim\Omega_k$ for small $\khat$.

Given the number of approximations made in \S\ref{adia_compare_gam1},
Fig. \ref{adia_growth_num} compares well with 
Fig. \ref{adia_growth}. The main qualitative features of the full
numerical solution in Fig. \ref{adia_growth_num} are captured by
Fig. \ref{adia_growth}: stabilization for 
$\khat\to0$ and $\khat\to\infty$,
stabilization for increasing $\gamma$ (but ineffective at small $\khat$), 
and the shift of the most unstable wavenumber to smaller values as
$\gamma$ is increased.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_real1}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag1}
  \caption{Real frequency (top) and growth rate 
    (bottom) of the fundamental VSI in a nearly vertically   
    isothermal disks ($\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$, 
    subject to adiabatic perturbations with different values of $\gamma$. 
    The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. This plot is to
    be compared with Fig. \ref{adia_growth}. \label{adia_growth_num}}   
\end{figure}

\subsection{Effect of thermal relaxation}
We now introduce thermal relaxation. Here we consider a slightly
different disk model with $\epsilon=0.05$ and $\gamma=1.4$, as
adopted in simulations performed by \cite{nelson13}. 
Fig. \ref{relax_growth_num} plots the
growth rates of the fundamental mode function of $\khat$.

%corrugation mode
% We first obtain the eigenvalues for the
% fundamental mode for $k_xH_\mathrm{iso}=0$, then vary parameters
% slowly to obtain eigenvalues for other values of $k_x$ or $\beta$.

Fig. \ref{relax_growth_num} is largely consistent with 
Fig. \ref{relax_disp_fig} except for $\khat\lesssim 1$
where the low-frequency approximation fails. 
For $\beta=0.063$ at $\khat\gtrsim50$ we found eigenfunctions had 
large perturbations near the vertical boundaries, which may be 
responsible for the non-smoothness in the curve at large $\khat$.

%likely attributable to the free surface boundary condition imposed in
%the numerical calculation (cf. regularity condition at infinity in
%\S\ref{disp_relax}).

Fig. \ref{relax_growth_num} also confirms our analytical discussion that
introducing a small but finite cooling time stabilizes the fundamental
VSI, but there is a preferable wavenumber that maximizes this effect (\S\ref{relax_pert}).
This occurs at $\khat= O(10)$ for the current disk model.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_imag_bloop}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag_bloop2}
  \caption{Growth rates of the fundamental VSI mode in 
    nearly-vertically isothermal disks ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$ and a range of thermal relaxation timescales 
    $\beta$. The eigenfrequencies are calculated by numerically 
    solving the full linear eigenvalue problem. \label{relax_growth_num}}   
\end{figure}

@@

\subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_vnoniso1} 
  \caption{Numerically-determined growth rates of the fundamental VSI
    mode in neutrally-buoyant ($N_z^2\equiv0$), vertically non-isothermal disks
    ($\Gamma=\gamma=1.4$), with $(p,q,\epsilon)=(0,-1,0.05)$, for
    a range of thermal relaxation timescales 
    $\beta$. \label{growth_vnoniso1}}     
\end{figure}

@@

\subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_vnoniso2}
  \caption{Same as Fig. \ref{growth_vnoniso1}, but with $\gamma=2$ so that
    $N_z^2\geq0$. \label{growth_vnoniso2}}
\end{figure}
